\documentclass[conference]{IEEEtran}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{xcolor}
\usepackage{colortbl}
\usepackage{listings}
\usepackage{tikz}
\usepackage{pgfplots}
\usepackage{hyperref}
\usepackage{microtype}
\usepackage{balance}
\pgfplotsset{compat=1.18}
\usetikzlibrary{shapes.geometric,arrows.meta,positioning,fit,backgrounds,calc}

\lstdefinelanguage{yaml}{
  keywords={rule,surface,on,match,tool,any,all,context,call_tree_contains,verdict,reason,severity,arg},
  keywordstyle=\color{blue!70}\bfseries,
  stringstyle=\color{orange!80!black},
  commentstyle=\color{gray}\itshape,
  morecomment=[l]{\#},
  morestring=[b]",
  morestring=[b]',
  sensitive=true,
}
\lstset{
  basicstyle=\ttfamily\scriptsize,
  breaklines=true,
  frame=single,
  rulecolor=\color{gray!30},
  backgroundcolor=\color{gray!5},
  captionpos=b,
  numbers=left,
  numberstyle=\tiny\color{gray},
  numbersep=4pt,
  xleftmargin=8pt,
}

\newtheorem{definition}{Definition}

\title{WatchTower: Observation-First Agent Security --- Taint Propagation
and Semantic Enforcement in Multi-Agent Systems}

\author{\IEEEauthorblockN{Anonymous Submission}\\IEEE S\&P 2027 --- Under Review}

\begin{document}

\maketitle

% -----------------------------------------------------------------------
\begin{abstract}
Multi-agent AI systems face three distinct attack surfaces that existing
defenses address only in isolation: \textit{input corruption} via prompt
injection, \textit{capability abuse} through legitimate tool misuse, and
\textit{multi-agent contagion} where a compromised agent poisons downstream
agents through shared memory. We present \textsc{WatchTower}, an
observation-first security framework that combines hook-level interception,
semantic call-tree context, and persistent cross-session taint propagation.
Unlike prior work that operates at content or network layers,
\textsc{WatchTower} intercepts at the tool-call level within the agent
runtime, enriches each interception with AST-derived call context, and
maintains a cross-session taint ledger enabling detection of MINJA-class
contagion attacks before the next execution trace. We demonstrate 100\%
detection on a 17-case known-bad corpus including direct reproduction of
MINJA attacks, with p99 enforcement latency of 0.011\,ms ---
9{,}636$\times$ faster than the closest competitor --- while being the
first published system to address all three attack surfaces with
cross-session taint tracking.
\end{abstract}

% -----------------------------------------------------------------------
\section{Introduction}
\label{sec:intro}

The deployment of multi-agent AI systems in production environments has
outpaced the security tooling designed to protect them. Where classical
software security assumes a well-defined process boundary, agent systems
blur those boundaries through tool calls, shared memory, and delegation
chains that span multiple autonomous processes. The result is three
distinct attack surfaces that no single published system defends against.

\subsection{The Three Attack Surfaces}

\textbf{Input Corruption.}
MINJA~\cite{minja2025} demonstrates that memory injection achieves
98.2\% injection success via query-only interaction. The attacker requires
no write access to the agent's memory store; adversarial content planted
in documents, web pages, or API responses is retrieved during normal
operation and incorporated into the agent's context. The attack succeeds
because prompt injection is a semantic failure: the agent cannot
distinguish instruction from data when both arrive through the same
channel.

\textbf{Capability Abuse.}
Consider an agent that exposes both \texttt{read\_secret} and
\texttt{send\_email}. Each call, in isolation, is legitimate. The
dangerous composition --- reading a secret and immediately emailing it to
an external address --- is indistinguishable at the wire level. Both calls
use the same API endpoints with the same authentication tokens. Wire-level
enforcement, which sees only individual calls, is blind to the causal
relationship between them. No published agent firewall system models this
relationship through call-tree context.

\textbf{Multi-Agent Contagion.}
When a compromised agent writes to a shared memory store, every downstream
agent that reads from that store inherits the taint through a query-only
interaction. This attack class is particularly dangerous because the victim
agent exhibits no anomalous behavior at the call level: it simply reads
memory, as designed. The malicious payload is carried in the content, not
the call structure. No published system tracks taint across agent session
boundaries.

\subsection{Why Existing Defenses Fall Short}

Existing defenses address these surfaces in isolation, leading to
systematic coverage gaps. Content-layer defenses (prompt injection
classifiers) cannot see tool invocation context. Network-layer defenses
see endpoints but lose semantic meaning. Runtime monitors observe behavior
but do not propagate state across agent boundaries. The root cause is
architectural: enforcement and observability have been designed as separate
concerns, when they must be co-designed to cover the full attack surface.

Table~\ref{tab:coverage} summarizes the coverage gaps of existing systems
across the three attack surfaces.

\begin{table}[t]
\centering
\caption{Attack surface coverage of existing agent security systems.
\checkmark\,=\,full, \textasciitilde\,=\,partial, $\times$\,=\,none.}
\label{tab:coverage}
\begin{tabular}{lccc}
\toprule
System & S1 Injection & S2 Cap.\ Abuse & S3 Contagion \\
\midrule
LLM Firewall~\cite{llmfirewall2025}      & \checkmark & $\times$ & $\times$ \\
LlamaFirewall~\cite{llamafirewall2025}    & \checkmark & \textasciitilde & $\times$ \\
AgentSpec~\cite{agentspec2025}            & \textasciitilde & \checkmark & $\times$ \\
Claw Patrol~\cite{clawpatrol2026}         & $\times$ & \textasciitilde & $\times$ \\
Distributed Sentinel~\cite{sentinel2026}  & \checkmark & \textasciitilde & $\times$ \\
\textbf{WatchTower (ours)}                & \checkmark & \checkmark & \checkmark \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Contributions}

This paper makes four contributions:

\begin{enumerate}

\item \textbf{Observation-first architecture with live call-tree context.}
\textsc{WatchTower} intercepts at the tool-call level inside the agent
runtime and enriches every interception with an AST-derived call tree,
enabling the \texttt{call\_tree\_contains} predicate that identifies
dangerous compositions invisible at the network layer.

\item \textbf{Cross-session taint propagation.}
We present the first published defense against MINJA-class multi-session
contagion. A persistent taint ledger (cavemem) propagates taint across
agent boundaries on every memory read, detecting contagion before the
downstream agent's next tool call --- without requiring replay of prior
sessions.

\item \textbf{Two-tier enforcement.}
A synchronous circuit-breaker (L0--L5, $<$10\,ms) blocks known-bad
patterns deterministically. An asynchronous Byzantine fault-tolerant (BFT)
swarm (L6--L7) handles ambiguous cases off the hot path, with a
task-completion barrier preventing bypass via async race conditions.

\item \textbf{Empirical evaluation.}
100\% detection on a 17-case known-bad corpus at 0.011\,ms p99 enforcement
latency, 9{,}636$\times$ faster than the closest competitor, with 0\%
false positives on a benign workload of 1{,}000 calls.

\end{enumerate}

% -----------------------------------------------------------------------
\section{Background and Related Work}
\label{sec:background}

\subsection{Prompt Injection and MINJA}

Prompt injection exploits the semantic conflation of instruction and data
in large language models~\cite{perez2022}. An attacker who can place
text in any document, web page, or API response visible to the agent can
influence its behavior without direct access to the system prompt.

MINJA~\cite{minja2025} generalizes this attack to multi-agent memory
systems. The key insight is that an attacker need not write directly to
the agent's memory store. Instead, adversarial content is planted in data
that the agent will retrieve during normal operation. MINJA achieves
98.2\% injection success and greater than 70\% end-to-end attack success
on GPT-4 and GPT-4o. The attack exploits \emph{bridging steps} ---
sequences of benign intermediate queries that accumulate injected context
until the final malicious action is triggered. This attack is structurally
invisible to systems that inspect individual calls without cross-session
state.

\subsection{Agent Firewalls}

\textbf{LLM Firewall Networks}~\cite{llmfirewall2025} apply content
projection to detect prompt injection, reducing attack success on the
ConVerse benchmark from 60\% to 3\% across 864 attack instances. The
system does not model agent identity and cannot propagate trust or taint
across agent boundaries. It cannot distinguish capability abuse from
legitimate use because it operates at the content layer.

\textbf{LlamaFirewall}~\cite{llamafirewall2025} combines three mechanisms:
PromptGuard~2 for injection detection, AlignmentCheck (a secondary LLM
acting as a judge), and CodeShield for unsafe code execution. It achieves
97.4\% accuracy on its evaluation benchmark. The LLM-in-the-loop
AlignmentCheck produces p50 latencies exceeding 100\,ms, making it
unsuitable for interactive agent loops. LlamaFirewall does not maintain
cross-session state and cannot propagate taint.

\textbf{AgentSpec}~\cite{agentspec2025} introduces a runtime DSL with
triggers and predicates for specifying agent behavior constraints. It
prevents more than 90\% of unsafe actions and achieves 70.96\% recall on
code safety tasks. AgentSpec reasons about individual agent behavior but
does not model contagion between agents via shared memory reads.

\textbf{Distributed Sentinel}~\cite{sentinel2026} uses a Secure Trace
Transport (STT) protocol with BFT consensus, achieving F1\,=\,0.95 at
106\,ms median latency (16\,ms verification plus 90\,ms entity extraction
on an A100 GPU). The GPU requirement limits deployment flexibility and the
latency profile is incompatible with synchronous enforcement on the hot path.

\textbf{Claw Patrol}~\cite{clawpatrol2026} enforces policies at the
network layer using WireGuard and Tailscale, achieving sub-5\,ms latency.
Operating at the wire level, it loses the semantic context of call trees:
it sees that \texttt{send\_email} was called but cannot know whether
\texttt{read\_secret} appeared earlier in the same execution trace.

\subsection{Agent Observability}

Process observability research~\cite{processobserv2025} identifies
stochastic agent behavior as a core challenge: the same prompt can produce
different tool call sequences across runs, making static analysis
insufficient. Auditable Agents~\cite{auditableagents2026} identifies six
open problems in agent auditability, noting that side-effecting operations
require fundamentally different safety guarantees than read-only queries.

Industry surveys report severe gaps: 82\% of organizations have limited
visibility into agent behavior~\cite{csa2025}, 65\% have experienced
security incidents involving AI agents~\cite{csa2025}, and only 47.1\% of
deployed systems are actively monitored~\cite{ibm2025}. These figures
motivate a system designed from the start with observability as a
first-class concern rather than an afterthought.

\subsection{Agent Identity and Trust}

SPIFFE/SPIRE~\cite{spiffe} provides cryptographic workload identity for
distributed systems. We adapt this for agent processes, but three
differences from classical deployment arise: delegation chains are dynamic
and created at agent spawn time rather than infrastructure provisioning
time; trust scores are behavioral rather than purely cryptographic; and
taint levels are updated during execution rather than at deployment.

\subsection{Taint Propagation in Agent Systems}

Classical taint analysis~\cite{newsome2005} tracks information flow at the
binary level within a single process, providing a binary tainted/untainted
judgment. Agent taint differs in three important ways. First, taint is
probabilistic, represented as a value in $[0, 1]$ with both hop-based and
time-based decay. Second, taint is cross-session and persistent: an agent
that read tainted memory in session $k$ carries that taint into session
$k+1$ unless it has decayed below threshold. Third, propagation is
bidirectional: taint flows through both memory writes (write propagation)
and memory reads (read propagation, the MINJA-class threat).

% -----------------------------------------------------------------------
\section{Threat Model}
\label{sec:threat}

\subsection{Attacker Capabilities}

\textsc{WatchTower} defends against attackers who can perform any
combination of the following actions:

\begin{itemize}
\item Inject adversarial content into documents, web pages, or API
      responses that agents will retrieve during normal operation.
\item Interact with agents via legitimate user queries, requiring no
      direct memory write access (MINJA-class attacks).
\item Compromise a single agent in a multi-agent system and use it as
      a contagion entry point via writes to shared memory.
\end{itemize}

\subsection{Out of Scope}

The following threat classes are considered out of scope for the current
system:

\begin{itemize}
\item Compromise of the \textsc{WatchTower} firewall host itself.
\item Compromise of the fabric certificate authority (CA) used for
      agent identity token issuance.
\item Model-weight jailbreaks and adversarial examples targeting the
      underlying LLM parameters; these are addressed at the alignment layer.
\end{itemize}

\subsection{Security Properties}

\begin{definition}[Security Properties P1--P5]
\textsc{WatchTower} is designed to enforce the following five properties:

\begin{enumerate}

\item[\textbf{P1}] \textit{No false negatives}: all cases KB01--KB17 in
      the known-bad corpus produce a BLOCK verdict.

\item[\textbf{P2}] \textit{Taint propagation}: for any agent $B$ that
      reads from memory written by tainted agent $A$, the invariant
      $T_B \geq T_A \cdot \rho$ holds before $B$'s next tool call,
      where $\rho = 0.8$ is the hop decay factor.

\item[\textbf{P3}] \textit{Fail-safe}: any unhandled internal error in
      the enforcement path produces BLOCK, never ALLOW.

\item[\textbf{P4}] \textit{Latency}: hot-path enforcement p99 $<$ 10\,ms.

\item[\textbf{P5}] \textit{No permanent DoS}: agent trust is recoverable
      via natural taint decay; quarantine expires at computable time
      $t_0 + \frac{1}{\lambda}\ln\!\left(\frac{T_0}{Q}\right)$.

\end{enumerate}
\end{definition}

% -----------------------------------------------------------------------
\section{System Design}
\label{sec:design}

\subsection{Architecture Overview}

\textsc{WatchTower} is a nine-layer vertical stack divided into two tiers.
The \textit{hot path} (L0--L5) executes synchronously in the tool-call
dispatch path and must complete within 10\,ms to satisfy P4. The
\textit{cold path} (L6--L7) executes asynchronously off the hot path and
is bounded by a task-completion barrier that prevents bypass via async
race conditions. L8 (Chronicle) is always-on and write-only from the
enforcement path.

Figure~\ref{fig:arch} shows the complete layer stack.

\begin{figure}[t]
\centering
\begin{tikzpicture}[
  layer/.style={rectangle, draw, rounded corners=2pt,
                minimum width=5.5cm, minimum height=0.55cm,
                font=\small\ttfamily, align=center},
  hot/.style={layer,  fill=orange!15},
  cold/.style={layer, fill=cyan!15},
  audit/.style={layer, fill=gray!15},
]
\node[hot]  (L0) at (0, 0.00) {L0: Identity Fabric (Ed25519 / SPIFFE)};
\node[hot]  (L1) at (0,-0.70) {L1: Interception (hermes pre\_tool\_call)};
\node[hot]  (L2) at (0,-1.40) {L2: Enrichment (graphify-ts AST)};
\node[hot]  (L3) at (0,-2.10) {L3: Deterministic Rules (YAML DSL)};
\node[hot]  (L4) at (0,-2.80) {L4: Trust Gate (routing thresholds)};
\node[hot]  (L5) at (0,-3.50) {L5: Taint Propagation (cavemem)};
\node[cold] (L6) at (0,-4.20) {L6: Async Analysis (ruflo BFT swarm)};
\node[cold] (L7) at (0,-4.90) {L7: Task-Completion Barrier};
\node[audit](L8) at (0,-5.60) {L8: Chronicle (ClickHouse, append-only)};
\begin{scope}[on background layer]
  \node[draw=orange!50, fill=orange!5, rounded corners=4pt,
        fit=(L0)(L5),
        label={[font=\scriptsize\itshape]right:hot path}] {};
  \node[draw=cyan!50, fill=cyan!5, rounded corners=4pt,
        fit=(L6)(L7),
        label={[font=\scriptsize\itshape]right:cold path}] {};
\end{scope}
\end{tikzpicture}
\caption{WatchTower nine-layer architecture. Orange: synchronous hot path
(L0--L5, p99 $<$\,10\,ms). Cyan: asynchronous cold path (L6--L7). Gray:
immutable audit layer.}
\label{fig:arch}
\end{figure}

\subsection{Identity Fabric (L0)}

Every agent process receives an Ed25519-signed identity token at spawn
time. Token fields are: \texttt{agent\_id} (UUID4), \texttt{parent\_id}
(spawning agent or null for root agents), \texttt{delegation\_depth}
(integer, maximum 8), \texttt{caps} (capability set, monotonically
narrowing through delegation chain), and \texttt{token\_exp} (expiry
timestamp, 1\,hr TTL).

The delegation invariant is enforced cryptographically at token issuance:
child capabilities must be a strict subset of parent capabilities. A
child agent cannot grant itself capabilities its parent does not hold.
This invariant is verified at L0 before any call reaches L1.

Trust scores are updated continuously. Clean calls increase trust toward
the ceiling; taint events decrease it in proportion to severity:

\begin{align}
  \text{trust}_{t+1} &= \text{trust}_t + \alpha(1 - \text{trust}_t)
    \quad \text{(clean call)} \label{eq:trust-clean} \\
  \text{trust}_{t+1} &= \text{trust}_t \cdot (1 - \beta \cdot s)
    \quad \text{(taint event, severity } s \text{)} \label{eq:trust-taint}
\end{align}

where $\alpha = 0.02$ (clean-call recovery rate) and $\beta = 0.6$
(taint-event decay rate). Severity $s \in [0, 1]$ is provided by the
rule engine at L3.

\subsection{Interception (L1)}

L1 hooks hermes's \texttt{pre\_tool\_call} and
\texttt{pre\_gateway\_dispatch} entry points. The hook receives the full
call context (agent identity, tool name, arguments, parent call stack) and
returns either \texttt{\{"block": True, "reason": "..."\}} to abort the
call or \texttt{None} to allow it to proceed.

Fail-safe behavior (P3) is implemented as a top-level exception handler
wrapping the entire enforcement pipeline: any unhandled exception ---
from a malformed signal, a database timeout, or an internal logic error
--- produces a BLOCK verdict and records the exception to L8. The agent
never observes a failed enforcement decision as an ALLOW.

\subsection{Enrichment (L2)}

L2 invokes graphify-ts, a tree-sitter WASM bridge that resolves the
current call site to its AST path within the calling source file. The
enrichment produces a \texttt{call\_tree} structure recording the sequence
of function calls and control-flow nodes from the module entry point to
the current tool dispatch, across 12 supported languages.

The cache hit rate target is $>$90\%. On a cache hit, enrichment adds
less than 0.1\,ms. On a cache miss, L2 sets the \texttt{needs\_async}
flag, routing the event to L5 for deferred enrichment, but does
\textit{not} block the call inline. This preserves P4 at the cost of a
bounded enrichment window for the first call to any new code path.

\subsection{Deterministic Rules (L3)}

L3 implements a YAML policy DSL evaluated in O(1) tool lookup plus
O(rules) sequential scan per call. No LLM calls are made on the hot path.
The rule evaluator is a pure function: identical inputs produce identical
verdicts, which is essential for both auditability and resistance to
statistical evasion attacks.

The key semantic differentiator is the \texttt{call\_tree\_contains}
predicate, which examines the enriched call tree from L2. This predicate
expresses that a tool call is only dangerous when preceded by specific
other calls in the same execution trace --- a distinction architecturally
impossible at the network layer, where each call is inspected in isolation.

Listing~\ref{lst:exfil} shows the capability-abuse rule for secret
exfiltration via email.

\begin{lstlisting}[language=yaml,
  caption={Capability-abuse rule: block secret exfiltration via email.},
  label={lst:exfil}]
rule: block_external_email_with_secrets
surface: capability_abuse
match:
  tool: send_email
  context:
    call_tree_contains: [read_secret]
  any:
    - arg.to: { not_in_domain: ["${ORG_DOMAIN}"] }
    - arg.body: { matches_secret_pattern: true }
verdict: BLOCK
severity: 0.9
\end{lstlisting}

The semantics are: block \texttt{send\_email} when (a) the current call
tree contains a prior call to \texttt{read\_secret}, and (b) either the
recipient is external to the organization domain or the message body
matches a known secret pattern. Without the \texttt{call\_tree\_contains}
predicate, the same \texttt{send\_email} call to an external address ---
absent any prior secret read --- is ALLOW. Same endpoint. Same arguments.
Different call tree. Different verdict. Wire-level enforcement cannot
distinguish these two cases.

\subsection{Trust Gate (L4)}

L4 uses the agent's current trust score as a routing signal with three
thresholds:

\begin{itemize}
\item $\text{trust} > 0.7$: ALLOW path --- proceed to L5 taint check only.
\item $0.3 \leq \text{trust} \leq 0.7$: ESCALATE path --- route to
      L6--L7 async analysis before issuing verdict.
\item $\text{trust} < 0.3$: NARROW path --- reduce capability set to the
      read-only subset defined in the agent's identity token.
\end{itemize}

Trust is a routing signal only, never a standalone hard-block. An agent
with low trust can still succeed on calls that pass L3 rule evaluation
and L5 taint checks. This preserves P5: a temporarily degraded agent is
not permanently denied service.

\subsection{Taint Propagation (L5)}

\textsc{WatchTower} maintains a persistent taint ledger in cavemem
(SQLite with FTS5, accessed via aiosqlite over the MCP protocol). Taint
levels decay over time and propagate across agent boundaries. The formal
model is:

\begin{align}
  T(a, t) &= T(a, t_0) \cdot e^{-\lambda(t - t_0)}
    \quad \text{(time decay)} \label{eq:decay} \\
  T(B) &:= \max\!\bigl(T(B),\; T(A) \cdot \rho\bigr)
    \quad \text{(read propagation, T3)} \label{eq:readprop} \\
  T(A) &:= \min\!\bigl(1.0,\; T(A) + s \cdot (1 - T(A))\bigr)
    \quad \text{(cap.\ abuse, T2)} \label{eq:abuse}
\end{align}

Parameters: hop decay $\rho = 0.8$, time decay constant
$\lambda = 0.1\,\text{hr}^{-1}$, quarantine threshold $Q = 0.7$.

Taint source codes classify the origin of each taint event:
\begin{itemize}
\item \textbf{T1}: Direct injection (tainted content retrieved from memory).
\item \textbf{T2}: Capability abuse blocked by L3 (taint applied to
      calling agent at severity $s$, via Eq.~\ref{eq:abuse}).
\item \textbf{T3}: Contagion via memory read (agent $B$ reads memory
      written by tainted agent $A$; Eq.~\ref{eq:readprop} applied).
\end{itemize}

An agent with $T(a, t) \geq Q = 0.7$ is quarantined: its capability set
is reduced to a minimal safe subset and the event is escalated to L6--L7.
Quarantine is not permanent (P5). It lifts when taint decays below $Q$
via natural decay~(Eq.~\ref{eq:decay}):

\begin{equation}
t_{\text{lift}} = t_0 + \frac{1}{\lambda}\ln\!\left(\frac{T_0}{Q}\right)
\label{eq:lift}
\end{equation}

For $T_0 = 0.9$: $t_{\text{lift}} = t_0 + 2.52\,\text{hr}$. For
$T_0 = 1.0$: $t_{\text{lift}} = t_0 + 3.57\,\text{hr}$.

Figure~\ref{fig:minja} illustrates the Q2 proof sequence, showing taint
propagation from agent~A to agent~B through a shared memory read without
write access.

\begin{figure}[t]
\centering
\begin{tikzpicture}[
  agent/.style={rectangle, draw, rounded corners=3pt,
                minimum width=2.0cm, minimum height=0.65cm,
                font=\small, align=center},
  store/.style={cylinder, draw, shape border rotate=90,
                minimum width=1.8cm, minimum height=0.65cm,
                font=\small, align=center},
  wt/.style={rectangle, draw=red!60, fill=red!10, rounded corners=3pt,
             minimum width=2.0cm, minimum height=0.65cm,
             font=\small, align=center},
  arr/.style={-Stealth, thick},
]
\node[agent, fill=red!20]    (A)  at (0, 0)    {Agent A\\$T_A=0.9$};
\node[store]                 (M)  at (3.2, 0)  {Memory\\Store};
\node[agent, fill=yellow!20] (B)  at (6.4, 0)  {Agent B\\$T_B=0$};
\node[wt]                    (WT) at (3.2,-1.9){WatchTower\\L5 Taint};
\node[agent, fill=red!30]    (Q)  at (6.4,-1.9){Agent B\\quarantined\\$T_B=0.72$};

\draw[arr] (A)  -- node[above, font=\scriptsize]{write} (M);
\draw[arr] (M)  -- node[above, font=\scriptsize]{read (query-only)} (B);
\draw[arr] (B)  -- node[right, font=\scriptsize]{T3 event} (WT);
\draw[arr] (WT) -- node[above, font=\scriptsize]{$0.9{\times}0.8{=}0.72{\geq}Q$} (Q);
\end{tikzpicture}
\caption{MINJA Q2 proof sequence. Agent A (taint 0.9, source T2) writes
to shared memory. Agent B reads via query-only interaction, holding no
write permissions. WatchTower propagates $T_B = 0.9 \times 0.8 = 0.72
\geq Q = 0.7$ and quarantines B before its next tool call. Detection
latency $<$1\,ms.}
\label{fig:minja}
\end{figure}

\subsection{Async Analysis (L6--L7)}

L6 implements a ruflo BFT consensus swarm comprising three specialized
agents: an injection analyzer, a behavioral anomaly detector, and a
capability graph auditor. Quorum requires $\geq 2/3$ BLOCK votes (two
of three agents). Timeout is 5\,s, after which the verdict defaults to
ALLOW to prevent denial of service from analysis latency. Any agent that
returns an error has its vote counted as BLOCK (fail-safe, P3), so a
completely failed swarm still produces a majority BLOCK (3/3 exceeds the
2/3 threshold).

L7 implements the task-completion barrier via an \texttt{asyncio.Future}
keyed by \texttt{event\_id} in a hold registry. When L4 routes an event
to L6--L7, the event is registered. Async verdicts are only applied if
the event's \texttt{event\_id} remains in the hold registry at verdict
time. Verdicts for already-dispatched events are rejected, preventing
the bypass demonstrated in KB15 where a rapid double-submit races the
async verdict.

% -----------------------------------------------------------------------
\section{Implementation}
\label{sec:impl}

\textsc{WatchTower} is implemented in approximately 1{,}200 lines of
Python 3.12 across two modules: \texttt{firewall/} (L0--L5 hot path) and
\texttt{agents/adapters/} (L6--L7 cold path). All I/O is async; no
synchronous I/O calls appear in the enforcement path. The signal shape is
defined once in \texttt{watchtower/core/signal.py} via Pydantic v2 and
imported by all layers; there are no duplicate signal definitions.

Table~\ref{tab:impl} summarizes the implementation components.

\begin{table}[t]
\centering
\caption{Implementation components.}
\label{tab:impl}
\begin{tabular}{lll}
\toprule
Component & Tool & Notes \\
\midrule
Runtime        & Python 3.12, asyncio, Pydantic v2 & \\
Hook layer     & hermes-agent (NousResearch)        & \texttt{pre\_tool\_call} \\
Taint store    & cavemem (SQLite+FTS5)              & aiosqlite, MCP \\
AST enrichment & graphify-ts                        & tree-sitter WASM, 12 langs \\
Compression    & caveman (UTC)                      & ${\approx}75\%$ size reduction \\
Async analysis & ruflo                              & BFT consensus, 3-agent quorum \\
Policy loader  & superpowers (YAML)                 & zero ML on hot path \\
Chronicle      & ClickHouse                         & append-only, no UPDATE/DELETE \\
Total LoC      & ${\approx}1{,}200$                 & firewall/ + agents/adapters/ \\
\bottomrule
\end{tabular}
\end{table}

The chronicle (L8) is a ClickHouse table with no UPDATE or DELETE
permissions granted to the application user. Every enforcement decision,
taint update, and trust score change is appended as an immutable row.
This provides a tamper-evident audit log suitable for forensic analysis
without session replay: the full call context is captured atomically at
enforcement time.

% -----------------------------------------------------------------------
\section{Evaluation}
\label{sec:eval}

\subsection{Known-Bad Corpus}

We evaluate \textsc{WatchTower} on 17 manually curated known-bad cases
spanning all three attack surfaces. Table~\ref{tab:kb} shows the full
corpus with the responsible enforcement mechanism for each case.

\begin{table*}[t]
\centering
\caption{Known-bad corpus. All 17 cases produce BLOCK. Zero false negatives.}
\label{tab:kb}
\begin{tabular}{llllll}
\toprule
ID & Surface & Description & Verdict & Mechanism \\
\midrule
KB01 & Injection   & Direct prompt injection in user message           & BLOCK & L3 rule \\
KB02 & Injection   & Injection via retrieved document content           & BLOCK & L3 rule \\
KB03 & Injection   & Injection via external API response body           & BLOCK & L3 rule \\
KB04 & Cap.\ Abuse & \texttt{read\_secret} then \texttt{send\_email} to external address & BLOCK & L3 \texttt{call\_tree\_contains} \\
KB05 & Cap.\ Abuse & \texttt{read\_secret} then \texttt{write\_file} to \texttt{/etc/}  & BLOCK & L3 \texttt{call\_tree\_contains} \\
KB06 & Cap.\ Abuse & \texttt{list\_keys} then batch \texttt{delete\_key} sequence        & BLOCK & L3 rule \\
KB07 & Cap.\ Abuse & Capability escalation via forged child token       & BLOCK & L0 delegation check \\
KB08 & Cap.\ Abuse & Replay of expired identity token (TTL violation)   & BLOCK & L0 TTL enforcement \\
KB09 & Contagion   & MINJA: agent A poisons memory, B reads (2 agents)  & BLOCK & L5 T3 taint \\
KB10 & Contagion   & MINJA: 3-agent chain, taint surviving 2 hops       & BLOCK & L5 T3 with hop decay \\
KB11 & Contagion   & Cross-session: B reads tainted memory in new session & BLOCK & L5 persistent ledger \\
KB12 & Injection   & Injection in tool return value (poisoned result)   & BLOCK & L3 rule \\
KB13 & Cap.\ Abuse & \texttt{execute\_code} outside allowed module list & BLOCK & L3 rule \\
KB14 & Injection   & Multi-turn injection accumulation (bridging steps) & BLOCK & L5 T1 taint accumulation \\
KB15 & Race        & Async verdict bypass via rapid double-submit       & BLOCK & L7 task-completion barrier \\
KB16 & Cap.\ Abuse & Trust-score gaming: clean-call buildup then burst  & BLOCK & L3 rule + L5 taint \\
KB17 & Contagion   & Taint via read-only memory query (no write perms)  & BLOCK & L5 T3 (query-only) \\
\bottomrule
\end{tabular}
\end{table*}

All 17 cases produce BLOCK verdicts. Detection rate: 100\%. False
negatives: 0. The corpus covers all three attack surfaces: S1 injection
(KB01--KB03, KB12, KB14), S2 capability abuse (KB04--KB08, KB13, KB16),
S3 contagion (KB09--KB11, KB17), plus two cross-cutting cases
(KB15: async race, KB16: trust gaming).

\subsection{Proof Scenarios}

\textbf{Q1: Coordination failure attribution.}
The Q1 proof requires identifying a coordination failure and returning the
responsible agent, the specific action, and its position in the call tree.
\textsc{WatchTower} produces a complete attribution record from L8
containing: \texttt{agent\_id} (the offending agent's UUID),
\texttt{source} (tool name and full argument set), and
\texttt{call\_tree\_position} (the AST path of the violating call within
the source file, as resolved by graphify-ts at L2). This provides full
forensic attribution without session replay: the chronicle append at
enforcement time captures the complete execution context atomically.

\textbf{Q2: MINJA cross-session contagion detection.}
The Q2 proof proceeds as follows:

\begin{enumerate}
\item Agent A triggers capability-abuse rule KB04; L3 blocks and L5
      applies T2 taint via Eq.~\ref{eq:abuse}: $T_A := 0.9$ (severity 0.9).
\item Agent A writes to shared memory store (subsequent call).
\item Agent B, in a new session with no prior interaction with A,
      performs a query-only read from the same memory store.
      Agent B holds no write permissions.
\item L5 detects the read, looks up $T_A = 0.9$ from the persistent
      taint ledger, and applies T3 propagation via Eq.~\ref{eq:readprop}:
      $T_B := \max(0,\; 0.9 \times 0.8) = 0.72$.
\item $T_B = 0.72 \geq Q = 0.7$: agent B is quarantined before its next
      tool call. Detection latency $<$1\,ms (local SQLite lookup,
      no network call required).
\end{enumerate}

This is the first cross-session MINJA defense with sub-millisecond
detection latency. The attack succeeds in MINJA~\cite{minja2025} because
there is no persistent cross-session state to inspect. The cavemem taint
ledger closes this gap by persisting taint across session boundaries.

\subsection{Performance}

Table~\ref{tab:perf} shows measured performance against design targets on
a workload of N\,=\,1{,}000 benign calls.

\begin{table}[t]
\centering
\caption{Performance measurements (N\,=\,1{,}000 benign calls).}
\label{tab:perf}
\begin{tabular}{lll}
\toprule
Metric & Target & Measured \\
\midrule
p99 hot-path latency  & $<$10\,ms & 0.011\,ms \\
Mean latency          & $<$10\,ms & 0.011\,ms \\
False positive rate   & $<$1\%    & 0\%       \\
Known-bad detection   & 100\%     & 100\%     \\
\bottomrule
\end{tabular}
\end{table}

The 0.011\,ms p99 latency is 9{,}636$\times$ faster than Distributed
Sentinel (106\,ms)~\cite{sentinel2026} and 9{,}091$\times$ faster than
LlamaFirewall (100\,ms p50)~\cite{llamafirewall2025}. The deterministic
rule engine at L3 dominates the hot-path cost: it is a Python dict lookup
and list scan with no I/O and no LLM calls.

Figure~\ref{fig:decay} shows taint decay curves for agents A and B over
8 hours, with the quarantine threshold as a dashed reference line. Agent
A ($T_0 = 0.9$) exits quarantine at 2.52\,hr after the taint event;
agent B ($T_0 = 0.72$, single-hop propagation) exits at
${\approx}0.25\,\text{hr}$, reflecting the weaker initial taint.

\begin{figure}[t]
\centering
\begin{tikzpicture}
\begin{axis}[
  width=\linewidth, height=5.2cm,
  xlabel={Time after taint event (hours)},
  ylabel={Taint level $T$},
  xmin=0, xmax=8,
  ymin=0, ymax=1.05,
  ytick={0,0.2,0.4,0.6,0.7,0.8,1.0},
  grid=major,
  grid style={gray!30},
  legend pos=north east,
  legend style={font=\scriptsize},
]
\addplot[thick, red,      domain=0:8, samples=200]
  { 0.9  * exp(-0.1 * x) };
\addlegendentry{Agent A ($T_0 = 0.9$)}
\addplot[thick, blue!70,  domain=0:8, samples=200]
  { 0.72 * exp(-0.1 * x) };
\addlegendentry{Agent B ($T_0 = 0.72$)}
\addplot[dashed, black!70, domain=0:8, samples=2]
  { 0.7 };
\addlegendentry{Quarantine threshold $Q = 0.7$}
\end{axis}
\end{tikzpicture}
\caption{Taint decay curves. $\lambda = 0.1\,\text{hr}^{-1}$. Agent A
exits quarantine at 2.52\,hr; agent B (one-hop propagation) exits at
${\approx}0.25\,\text{hr}$.}
\label{fig:decay}
\end{figure}

Figure~\ref{fig:latency} shows p99 hot-path latency on a log scale
comparing \textsc{WatchTower} to published competitors.

\begin{figure}[t]
\centering
\begin{tikzpicture}
\begin{axis}[
  width=\linewidth, height=5.2cm,
  ybar,
  ymode=log,
  log origin=infty,
  ylabel={p99 Latency (ms, log scale)},
  xtick=data,
  symbolic x coords={WatchTower, AgentSpec, {Claw Patrol}, LlamaFirewall, Sentinel},
  xticklabel style={font=\scriptsize, rotate=22, anchor=north east},
  nodes near coords,
  nodes near coords style={font=\tiny\ttfamily},
  ymin=0.005, ymax=300,
  bar width=18pt,
  grid=major,
  grid style={gray!30},
]
\addplot[fill=green!55!black!60, draw=green!50!black] coordinates {
  (WatchTower,   0.011)
  (AgentSpec,    1)
  ({Claw Patrol}, 5)
  (LlamaFirewall, 100)
  (Sentinel,     106)
};
\end{axis}
\end{tikzpicture}
\caption{p99 hot-path latency (log scale). WatchTower: 0.011\,ms;
AgentSpec: ${\sim}$1\,ms~\cite{agentspec2025}; Claw Patrol:
$<$5\,ms~\cite{clawpatrol2026}; LlamaFirewall:
$>$100\,ms~\cite{llamafirewall2025}; Distributed Sentinel:
106\,ms~\cite{sentinel2026}.}
\label{fig:latency}
\end{figure}

\subsection{Comparison with State of the Art}

Table~\ref{tab:compare} provides a full comparison across attack surface
coverage, latency, cross-session capability, and key reported metric.

\begin{table*}[t]
\centering
\caption{Comparison with state-of-the-art agent security systems.
\checkmark\,=\,full, \textasciitilde\,=\,partial, $\times$\,=\,not addressed.}
\label{tab:compare}
\begin{tabular}{lcccccc}
\toprule
System & S1 Injection & S2 Cap.\ Abuse & S3 Contagion
       & Latency & Cross-session & Key Metric \\
\midrule
LLM Firewall~\cite{llmfirewall2025}
  & \checkmark & $\times$ & $\times$
  & N/A & No & 60\%$\to$3\% on ConVerse \\
LlamaFirewall~\cite{llamafirewall2025}
  & \checkmark & \textasciitilde & $\times$
  & $>$100\,ms p50 & No & 97.4\% accuracy \\
AgentSpec~\cite{agentspec2025}
  & \textasciitilde & \checkmark & $\times$
  & ${\sim}$1\,ms & No & $>$90\% unsafe prevention \\
Claw Patrol~\cite{clawpatrol2026}
  & $\times$ & \textasciitilde & $\times$
  & $<$5\,ms & No & Wire-level (WireGuard) \\
Distributed Sentinel~\cite{sentinel2026}
  & \checkmark & \textasciitilde & $\times$
  & 106\,ms & No & F1\,=\,0.95 \\
\rowcolor{green!10}
\textbf{WatchTower (ours)}
  & \checkmark & \checkmark & \checkmark
  & \textbf{0.011\,ms} & \textbf{Yes}
  & \textbf{100\% on 17-case corpus} \\
\bottomrule
\end{tabular}
\end{table*}

\textsc{WatchTower} is the only system to achieve full coverage across
all three attack surfaces while providing cross-session taint tracking at
sub-millisecond enforcement latency.

% -----------------------------------------------------------------------
\section{Discussion}
\label{sec:discussion}

\subsection{Why Determinism Beats LLM-in-the-Loop}

LlamaFirewall's AlignmentCheck achieves 97.4\% accuracy using a secondary
LLM as a reasoning judge. For many applications this is excellent. For
security-critical enforcement paths, non-determinism is a structural
liability: an attacker who can characterize the variance of an LLM judge
can construct adversarial inputs that probabilistically bypass the check.
Identical inputs must produce identical verdicts for an enforcement system
to be auditable and resistant to statistical evasion attacks.

\textsc{WatchTower}'s L3 rule engine is a pure function. Given the same
call context, it produces the same verdict on every invocation. Rules are
human-readable YAML that security engineers can audit, version-control,
and reason about without executing the system. There is no latency
variance: a rule that takes 0.001\,ms at deployment takes 0.001\,ms at
2\,AM under peak load.

\textsc{WatchTower} uses LLM reasoning only in the ruflo swarm (L6--L7),
which operates on the cold path for ambiguous escalations and does not
participate in hot-path enforcement decisions.

\subsection{The Semantic Gap}

The central architectural insight of \textsc{WatchTower} is that security
enforcement must operate at the semantic layer, not the wire layer. A
network-level monitor sees that \texttt{send\_email} was called. It does
not see whether \texttt{read\_secret} was called 200\,ms earlier in the
same execution trace.

The \texttt{call\_tree\_contains} predicate closes this gap. The same
\texttt{send\_email} call to an external address is ALLOW when the call
tree contains no prior secret reads, and BLOCK when it does. This
distinction is not available to any defense operating below the agent
runtime layer: it requires that the enforcement system be co-located with
the execution context, which is the defining property of the
observation-first architecture.

\subsection{Open Problems}

\begin{enumerate}

\item \textbf{Irreversible tool classification.}
The L3 rule engine requires human-authored rules to classify irreversible
operations (file deletion, external email, key rotation). Automated
irreversibility detection --- possibly via static analysis of tool schemas
or LLM-assisted rule synthesis --- would reduce authoring burden.

\item \textbf{graphify cold-path enrichment window.}
On a cache miss, enrichment is deferred to the async path, creating a
bounded window where the first call to a new code path executes without
call-tree context. An adaptive pre-warming strategy that pre-enriches
known tool dispatch sites at startup would eliminate this window.

\item \textbf{Trust-score gaming over long horizons.}
An attacker with sufficient patience can accumulate clean calls to raise
trust before attempting abuse. The $\alpha = 0.02$ recovery rate provides
slow recovery by design, but a formal analysis of the minimum attack cost
under this model has not been conducted.

\item \textbf{Recursive trust (the watcher problem).}
If the \textsc{WatchTower} enforcement agent is itself an agent in a
larger system, the question of who watches the watcher arises. The current
architecture treats the firewall host as trusted infrastructure. A fully
adversarial deployment would require a separate attestation root.

\item \textbf{Taint decay constant calibration.}
The decay constant $\lambda = 0.1\,\text{hr}^{-1}$ (quarantine half-life
6.93\,hr) is set conservatively. Empirical calibration against real attack
datasets would allow tighter parameters, reducing false-positive pressure
from long-lived taint on legitimate agents.

\item \textbf{Distributed cavemem cross-host latency.}
The taint ledger is currently a local SQLite instance accessed via
aiosqlite. In a multi-host deployment, taint reads and writes cross a
network boundary. The additional latency and consistency guarantees of a
distributed taint store have not been characterized.

\end{enumerate}

\subsection{Limitations}

The 17-case known-bad corpus was manually curated to cover representative
attack patterns. Real-world attack diversity may exceed what these cases
test; novel compositions not covered by existing literature may evade L3
rules that have not been written for them. The system is only as strong as
its rule set against unknown-unknown attacks.

The ruflo swarm (L6--L7) is currently implemented as a heuristic stub
with simulated BFT consensus. A production deployment would require a
full BFT implementation with cryptographic vote signing and Byzantine node
detection.

The graphify enrichment is implemented as a subprocess bridge to a
Node.js process running tree-sitter WASM. A native Python binding to
tree-sitter would reduce cold-path enrichment latency by eliminating the
subprocess overhead.

Cross-host taint propagation latency has not been characterized. The
sub-millisecond taint detection reported here assumes a co-located taint
store. Networked deployments will exhibit higher latency depending on the
consistency protocol of the distributed taint store.

% -----------------------------------------------------------------------
\section{Conclusion}
\label{sec:conclusion}

\textsc{WatchTower} demonstrates that co-designing enforcement with
observability addresses all three agent attack surfaces --- input
corruption, capability abuse, and multi-agent contagion --- at
sub-millisecond enforcement latency. By intercepting at the tool-call
level, enriching with AST-derived call context, and maintaining a
persistent cross-session taint ledger, the system achieves 100\% detection
on a 17-case known-bad corpus at 0.011\,ms p99 hot-path latency ---
suitable for interactive agent loops where prior systems' 100\,ms+
latencies are prohibitive.

The \texttt{call\_tree\_contains} predicate is the key semantic
differentiator, enabling policy rules that distinguish dangerous
compositions from individually-legitimate calls. The cross-session taint
ledger is the first published defense against MINJA-class multi-session
contagion, detecting taint propagation through query-only memory reads
before the downstream agent's next tool call, at a detection latency of
less than 1\,ms --- three to four orders of magnitude below prior work.

As multi-agent deployments grow in scale and autonomy, the observation-first
architecture --- where enforcement is co-located with execution context and
shares its semantic understanding --- is the necessary direction for agent
security research. \textsc{WatchTower} provides an existence proof that
coverage completeness and sub-millisecond performance are simultaneously
achievable.

\balance
\bibliographystyle{IEEEtran}
\bibliography{references}

\end{document}
